Towards Practical First-Order Model Counting

### 2024-04-24 doc/paper.tex
Section 6

\section{Experimental Evaluation (TODO)}

We compare \Cranetwo{} (in both BFS and greedy modes) with \textsc{FastWFOMC}
and \textsc{ForcLift} on two problems previously considered by Dilkas and
Belle~\shortcite{DBLP:conf/kr/DilkasB23}: the function-counting problem from
\cref{example:functions} and the bijection-counting problem described below.
Note that comparing \Cranetwo{} and \textsc{FastWFOMC} on a larger set of
benchmarks is challenging because there is no automated way to translate a
formula in \FO{} or \Ctwo{} into \UFO{} (or even check if such an encoding is
possible).

@@

We run each of the four algorithms once on each combination of formula and
domain size, the latter ranging from one up to and including 35. The experiments
were run on an AMD~Ryzen~7~5800H processor with \SI{16}{\gibi\byte} of memory
and Arch Linux~6.8.2-arch2-1 operating system. \textsc{FastWFOMC} was executed
using Python~3.8.19 with Python-FLINT~0.5.0.

@@

\Cref{fig:plot} shows total runtime values as a measure of the overall
performance of each algorithm. For the three knowledge compilation algorithms,
we also separately track compilation and inference times and comment on them
where relevant. Note that \Cranebfs{} is able to handle more instances than
either \textsc{ForcLift} or \Cranegreedy{}, i.e., the bijection-counting problem
in our experiments as well as similar formulas examined
previously~\cite{DBLP:conf/kr/DilkasB23}.

As shown in the figure, the runtimes of all compilation-based algorithms remain
practically constant in contrast to the rapidly increasing runtimes of
\textsc{FastWFOMC}. Although the search/compilation part is slower in
\Cranetwo{} than in \textsc{ForcLift}, the difference is negligible. The
runtimes of the knowledge compilation algorithms appear constant because---for
these counting problems and domain sizes---compilation time dominates inference
time (recall that compilation time is independent of domain sizes). Indeed, the
maximum inference time of both \Cranebfs{} and \Cranegreedy{} across these
experiments is only \SI{4}{\milli\second}. The runtimes of \Cranetwo{} have
lower variation than those of \textsc{ForcLift} because with \textsc{ForcLift}
we compile the formula anew for each domain size whereas with \Cranetwo{} we
compile it once and reuse the resulting C++ program for all domain sizes.

As another point of comparison, we examine the maximum domain size that the
algorithms can handle in at most \SI{45}{\second} of total runtime. For
bijection-counting, \textsc{FastWFOMC} scales up to domains of size 28, whereas
\Cranebfs{} can handle domains of size \num{e4}. In the case of
function-counting, \textsc{FastWFOMC} can handle domains of size 32, whereas
\Cranebfs{} can handle \num{3e5}. On function-counting, \textsc{ForcLift} can
handle domains sizes up to about 100, but then the finite-precision arithmetic
declares higher model as infinite.

\todo[inline,caption={}]{
  \begin{itemize}
    \item Add info about \textsc{ForcLift} (taking care of infinities)
    \item Turn the above paragraph into a table.
    \item Mention two types of experiments earlier in the section.
  \end{itemize}
}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Later on:
          \begin{itemize}
            \item Eliminate (e.g., Emacs) warnings.
            \item must run it by all 3 other coauthors
            \item re-check submission instructions and formatting guidelines
            \item Full stop at the end of a single-sentence caption?
            \item cite the \textsc{Crane} paper (and other algorithms' papers)
                  where necessary
            \item Something to think about that could inspire some kind of
                  theorems: how can I assume that a domain is empty if there are
                  constants associated with it?
            \item consider turning small sections (e.g., Section 5) into
                  subsections of even paragraphs.
            \item Authors may submit a separate PDF with additional information
                  supporting their claims (such as proof details, additional
                  experimental results, further details on experimental design,
                  etc).
          \end{itemize}
  \end{itemize}
}

>>>>

\section{Experimental Evaluation}

We compare \Cranetwo{} (in both BFS and greedy modes) with \textsc{FastWFOMC}
and \textsc{ForcLift} on two problems previously considered by Dilkas and
Belle~\shortcite{DBLP:conf/kr/DilkasB23}: the function-counting problem from
\cref{example:functions} and the bijection-counting problem described below.
Note that comparing \Cranetwo{} and \textsc{FastWFOMC} on a larger set of
benchmarks is challenging because there is no automated way to translate a
formula in \FO{} or \Ctwo{} into \UFO{} (or even check if such an encoding is
possible). The experiments were run on an AMD~Ryzen~7~5800H processor with
\SI{16}{\gibi\byte} of memory and Arch Linux~6.8.2-arch2-1 operating system.
\textsc{FastWFOMC} was executed using Python~3.8.19 with Python-FLINT~0.5.0.

@@

@@

In the first experiment, we run each of the four algorithms once on each
combination of benchmark and domain size, the latter ranging from one up to and
including 35. \Cref{fig:plot} shows total runtime values as a measure of the
overall performance of each algorithm. For the three knowledge compilation
algorithms, we also separately track compilation and inference times and comment
on them where relevant. Note that \Cranebfs{} is able to handle more instances
than either \textsc{ForcLift} or \Cranegreedy{}, i.e., the bijection-counting
problem in our experiments as well as similar formulas examined
previously~\cite{DBLP:conf/kr/DilkasB23}.

As shown in \cref{fig:plot}, the runtimes of all compilation-based algorithms
remain practically constant in contrast to the rapidly increasing runtimes of
\textsc{FastWFOMC}. Although the search/compilation part is slower in
\Cranetwo{} than in \textsc{ForcLift}, the difference is negligible. The
runtimes of the knowledge compilation algorithms appear constant because---for
these counting problems and domain sizes---compilation time dominates inference
time (recall that compilation time is independent of domain sizes). Indeed, the
maximum inference time of both \Cranebfs{} and \Cranegreedy{} across these
experiments is only \SI{4}{\milli\second}. The runtimes of \Cranetwo{} have
lower variation than those of \textsc{ForcLift} because with \textsc{ForcLift}
we compile the formula anew for each domain size whereas with \Cranetwo{} we
compile it once and reuse the resulting C++ program for all domain sizes.

\begin{table}
  \centering
  \begin{tabular}{lrr}
    \toprule
    Algorithm & Bijections & Functions \\
    \midrule
    \Cranebfs{} & \num{e4} & \num{3e5} \\
    \Cranegreedy{} & --- & \num{3e5} \\
    \textsc{FastWFOMC} & 28 & 32 \\
    \textsc{ForcLift} & --- & 143 \\
    \bottomrule
  \end{tabular}
  \caption{The maximum domain sizes that can be handled by each algorithm in at
    most \SI{45}{\second}. `---' indicates that the algorithm is unable to find
    a solution. The domain sizes for \Cranetwo{} are accurate up to the first
    digit while other domain sizes are exact.}\label{table:results}
\end{table}

For the second experiment, we examine the maximum domain size that the
algorithms can handle in at most \SI{45}{\second} of total runtime. As shown in
\cref{table:results}, \Cranebfs{} is able to scale to domain sizes \num{357} and
\num{9375} times larger than \textsc{FastWFOMC} for bijection-counting and
function-counting problems, respectively. Note that, while \textsc{ForcLift} may
have runtime performance comparable to \Cranetwo{}, its finite-precision
arithmetic cannot represent model counts for domain sizes higher than 143 and
returns $\infty$ instead.

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Later on:
          \begin{itemize}
            \item Eliminate (e.g., Emacs) warnings.
            \item must run it by all 3 other coauthors
            \item re-check submission instructions and formatting guidelines
            \item cite the \textsc{Crane} paper (and other algorithms' papers)
                  where necessary
          \end{itemize}
  \end{itemize}
}
